Some examples in extended logic counting classes

### 2016-08-03 QSO/cap-comp.tex
Nueva versión de QSO.

Assume that $\SR = (S, \adds, \mults, \zero, \one)$ is a semiring and $\theta$ is a sentence in $\qso(\SR)$. Then slightly abusing notation, we also use $\theta$ to denote a function from $\R$ to $\SR$ such that for every $\A \in \str[\R] $:
\begin{eqnarray*}
\theta(\enc(\A)) & = & \sem{\theta}(\A).
\end{eqnarray*}
\begin{definition}
Let $\SR$ be a semiring, $\FF$ a fragment of $\qso(\SR)$, $\CC$ a function complexity class over $\SR$ and $\KK$ a class of finite structures. Then $\FF$ {\em captures} $\CC$ over $\KK$ if:
\begin{itemize}
\item for every $\theta$ in $\FF$, there exists $f \in \CC$ such that $\res{\theta}{\KK} = \res{f}{\KK}$; and

\item for every $f \in \CC$, there exists $\theta$ in $\FF$ such that $\res{f}{\KK} = \res{\theta}{\KK}$.
\end{itemize}
\end{definition}
In the previous definition, if $\KK$ is the class of all finite structures, then we say that $\FF$ captures $\CC$ over the class of all structures, and if $\KK$is the class of all ordered finite structures, then we say that $\FF$ captures $\CC$ over the class of ordered structures. 
\begin{theorem}
$\eqso(\nat,\fo)$ captures $\shp$ over the class of ordered structures.
\end{theorem}

\begin{proposition}
$\eqso(\nat,\fo)$ does not capture $\shp$ over the class of all structures.
\end{proposition}

\begin{theorem}
$\eqso(\integ,\fo)$ captures $\gp$ over the class of ordered structures.
\end{theorem}

\begin{theorem}
$\eqso(\nat,\eso)$ captures $\spp$ over the class of ordered structures.
\end{theorem}

>>>>

Assume that $\alpha$ is a sentence in $\qso$. Then slightly abusing notation, we also use $\alpha$ to denote a function from $\R$ to $\bbN$ such that for every $\A \in \str[\R] $:
\begin{eqnarray*}
\alpha(\enc(\A)) & = & \sem{\alpha}(\A).
\end{eqnarray*}
\begin{definition}
Let $\FF$ a fragment of $\qso$, $\CC$ a function complexity class over $\bbN$ and $\KK$ a class of finite structures. Then $\FF$ {\em captures} $\CC$ over $\KK$ if:
\begin{itemize}
\item for every $\alpha$ in $\FF$, there exists $f \in \CC$ such that $\res{\alpha}{\KK} = \res{f}{\KK}$; and

\item for every $f \in \CC$, there exists $\alpha$ in $\FF$ such that $\res{f}{\KK} = \res{\alpha}{\KK}$.
\end{itemize}
\end{definition}
In the previous definition, if $\KK$ is the class of all finite structures, then we say that $\FF$ captures $\CC$ over the class of all structures, and if $\KK$is the class of all ordered finite structures, then we say that $\FF$ captures $\CC$ over the class of ordered structures.

\cristian{Aca hay que definir esto sobre estructuras finitas directamente. No veo el sentido de tener estructuras infinitas.}

\begin{theorem}
$\eqso(\fo)$ captures $\shp$ over the class of ordered structures.
\end{theorem}

\begin{proposition}
$\eqso(\fo)$ does not capture $\shp$ over the class of all structures.
\end{proposition}

%\begin{theorem}
%$\eqso(\integ,\fo)$ captures $\gp$ over the class of ordered structures.
%\end{theorem}

\begin{theorem}
$\eqso(\eso)$ captures $\spp$ over the class of ordered structures.
\end{theorem}

### 2016-08-03 QSO/logics.tex
Nueva versión de QSO.

Fix a semiring $\SR = (S, \adds, \mults, \zero, \one)$. Then given a relational signature $\R$, the set of $\SR$-quantitative second-order logic formulas (or just $\qso(\SR)$-formulas) over $\R$ is given by the following grammar:
\begin{eqnarray}
\label{eq-def-qso}
  \theta & := & \varphi \ \mid \ s \ \mid \ (\theta \add \theta) \ \mid\ (\theta \mult \theta) \ \mid \ \sa{x} \theta \ \mid \ \pa{x} \theta \ \mid \ \sa{X} \theta \ \mid \ \pa{X} \theta 
\end{eqnarray}
where $\varphi$ is an $\so$-formula over $\R$, $s \in S$, $x \in \fv$ and $X \in \sv$.

Let $\R$ be a relational signature, $\A$ be a finite $\R$-structure with domain $A$, $v$ a first-order variable assignment for $\A$ and $V$ a second-order variable assignment for $\A$. Then the \emph{evaluation} of a $\qso(\SR)$-formula $\theta$ over $(\A, v, V)$ is defined as a function $\sem{\theta}$ that on input $(\A, v, V)$ returns a value in the semiring $\SR$. Formally, the function $\sem{\theta}$ is recursively defined as follows:
\begin{eqnarray*}
\sem{\varphi}(\A, v, V) & = & 
\begin{cases}
\one & \mbox{if } (\A, v, V) \models \varphi \\
\zero & \mbox{otherwise}
\end{cases}\\
\sem{s}(\A, v, V) & = & s \\
\sem{\theta_1 \add \theta_2}(\A, v, V) & = & \sem{\theta_1}(\A, v, V) \adds \sem{\theta_2}(\A, v, V)\\
\sem{\theta_1 \mult \theta_2}(\A, v, V) & = & \sem{\theta_1}(\A, v, V) \mults \sem{\theta_2}(\A, v, V)\\ 
\sem{\sa{x} \theta}(\A, v, V) & = & \bigadds_{a \in A} \sem{\theta}(\A,v[a/x],V)\\
\sem{\pa{x} \theta}(\A, v, V) & = & \bigmults_{a \in A} \sem{\theta}(\A,v[a/x],V)\\
\sem{\sa{X} \theta}(\A, v, V) & = & \bigadds_{B \subseteq A^{\arity(X)}} \sem{\theta}(\A, v, V[B/X])\\
\sem{\pa{X} \theta}(\A, v, V) & = & \bigmults_{B \subseteq A^{\arity(X)}} \sem{\theta}(\A, v, V[B/X])
\end{eqnarray*}
A formula in $\qso(\SR)$ can mention the usual quantifiers in $\so$ (that is, $\exists x$ and $\exists X$) and the quantifiers that make use of the two operations in the semiring (that is, $\Sigma x$, $\Pi x$, $\Sigma X$ and $\Pi X$), which are called {\em semiring quantifiers} . A $\qso(\SR)$-formula $\theta$ is said to be a \emph{sentence} if it does not have any free variable, that is, every variable in $\theta$ is under the scope of a usual quantifier or a semiring quantifier. It is important to notice that if $\theta$ is a $\qso(\SR)$-sentence over a relational signature $\R$, then for every finite $\R$-structure $\A$, first-order variable assignments $v_1$, $v_2$ for $\A$ and second-order variable assignments $V_1$, $V_2$ for $\A$, it holds that:
\begin{eqnarray*}
\sem{\theta}(\A, v_1, V_1) & = & \sem{\theta}(\A, v_2, V_2).
\end{eqnarray*}
Thus, in such a case we use the term $\sem{\theta}(\A)$ to denote $\sem{\theta}(\A, v, V)$, for some arbitrary first-order variable assignment $v$ for $\A$ and some arbitrary second-order variable assignment $V$ for $\A$.

In this paper, we consider several fragments of $\qso(\SR)$, which are obtained by restricting the syntax of the formula $\varphi$ in \eqref{eq-def-qso} or the use of the semiring quantifiers. Let $\eqso(\SR)$ be the fragment of $\qso(\SR)$ that is obtained by only allowing the semiring quantifiers $\Sigma x$, $\Pi x$ and $\Sigma X$ in \eqref{eq-def-qso}. Moreover, assuming that $\LL$ is a fragment of $\so$, let $\qso(\SR, \LL)$ be the fragment of $\qso(\SR)$ obtained by restricting $\varphi$ in \eqref{eq-def-qso} to be a formula in $\LL$, and let $\eqso(\SR, \LL)$ be the fragment of $\eqso(\SR)$ obtained by imposing the same restriction. In particular, in this paper we consider the following fragments $\LL$: $\fo$, which is obtained by disallowing the use of the second-order quantifier ($\exists X$) in the formula $\varphi$ in \eqref{eq-def-qso}, and $\eso$, which is obtained by allowing $\varphi$ in \eqref{eq-def-qso} to be a formula of the form $\exists X_1 \cdots \exists X_k \, \psi$, where $\psi$ is an $\fo$-formula.

>>>>

Consider $\bbN$ as the set of non-negative integers. Then given a relational signature $\R$, the set of quantitative second-order logic formulas (or just $\qso$-formulas) over $\R$ is given by the following grammar:
\begin{eqnarray}
\label{eq-def-qso}
  \alpha & := & \varphi \ \mid \ s \ \mid \ (\alpha \add \alpha) \ \mid\ (\alpha \mult \alpha) \ \mid \ \sa{x} \alpha \ \mid \ \pa{x} \alpha \ \mid \ \sa{X} \alpha \ \mid \ \pa{X} \alpha 
\end{eqnarray}
where $\varphi$ is an $\so$-formula over $\R$, $s \in \bbN$, $x \in \fv$ and $X \in \sv$.

Let $\R$ be a relational signature, $\A$ be a finite $\R$-structure with domain $A$, $v$ a first-order assignment for $\A$ and $V$ a second-order assignment for $\A$. Then the \emph{evaluation} of a $\qso$-formula $\alpha$ over $(\A, v, V)$ is defined as a function $\sem{\alpha}$ that on input $(\A, v, V)$ returns a non-negative integer in $\bbN$. Formally, the function $\sem{\alpha}$ is recursively defined as follows:
$$
\renewcommand{\arraystretch}{1.7}
\begin{array}{rcl} 
\sem{\varphi}(\A, v, V) & = & 
\begin{cases}
1 & \mbox{if } (\A, v, V) \models \varphi \\
0 & \mbox{otherwise}
\end{cases}\\
\sem{s}(\A, v, V) & = & s \\
\sem{\alpha_1 \add \alpha_2}(\A, v, V) & = & \sem{\alpha_1}(\A, v, V) + \sem{\alpha_2}(\A, v, V)\\
\sem{\alpha_1 \mult \alpha_2}(\A, v, V) & = & \sem{\alpha_1}(\A, v, V) \cdot \sem{\alpha_2}(\A, v, V)\\ 
\sem{\sa{x} \alpha}(\A, v, V) & = & \displaystyle \sum_{a \in A} \sem{\alpha}(\A,v[a/x],V)\\
\sem{\pa{x} \alpha}(\A, v, V) & = & \displaystyle \prod_{a \in A} \sem{\alpha}(\A,v[a/x],V)\\
\sem{\sa{X} \alpha}(\A, v, V) & = & \displaystyle \sum_{B \subseteq A^{\arity(X)}} \sem{\alpha}(\A, v, V[B/X])\\
\sem{\pa{X} \alpha}(\A, v, V) & = & \displaystyle \prod_{B \subseteq A^{\arity(X)}} \sem{\alpha}(\A, v, V[B/X])
\end{array}
$$
A formula in $\qso$ can mention the usual quantifiers in $\so$ (that is, $\exists x$ and $\exists X$) and the quantifiers that make use of addition and multiplication (that is, $\Sigma x$, $\Pi x$, $\Sigma X$ and $\Pi X$), which are called {\em quantitative quantifiers} . A $\qso$-formula $\alpha$ is said to be a \emph{sentence} if it does not have any free variable, that is, every variable in $\alpha$ is under the scope of a usual quantifier or a quantitative quantifier. It is important to notice that if $\alpha$ is a $\qso$-sentence over a relational signature $\R$, then for every finite $\R$-structure $\A$, first-order assignments $v_1$, $v_2$ for $\A$ and second-order assignments $V_1$, $V_2$ for $\A$, it holds that:
\begin{eqnarray*}
\sem{\alpha}(\A, v_1, V_1) & = & \sem{\alpha}(\A, v_2, V_2).
\end{eqnarray*}
Thus, in such a case we use the term $\sem{\alpha}(\A)$ to denote $\sem{\alpha}(\A, v, V)$, for some arbitrary first-order assignment $v$ for $\A$ and some arbitrary second-order assignment $V$ for $\A$.

In this paper, we consider several fragments of $\qso$, which are obtained by restricting the syntax of the formula $\varphi$ in \eqref{eq-def-qso} or the use of the quantitative quantifiers. Let $\eqso$ be the fragment of $\qso$ that is obtained by only allowing the quantitative quantifiers $\Sigma x$, $\Pi x$ and $\Sigma X$ in \eqref{eq-def-qso}. Moreover, assuming that $\LL$ is a fragment of $\so$, let $\qso(\LL)$ be the fragment of $\qso$ obtained by restricting $\varphi$ in \eqref{eq-def-qso} to be a formula in $\LL$, and let $\eqso(\LL)$ be the fragment of $\eqso$ obtained by imposing the same restriction. In particular, in this paper we consider the following fragments $\LL$: $\fo$, which is obtained by disallowing the use of the second-order quantifier ($\exists X$) in the formula $\varphi$ in \eqref{eq-def-qso}, and $\eso$, which is obtained by allowing $\varphi$ in \eqref{eq-def-qso} to be a formula of the form $\exists X_1 \cdots \exists X_k \, \psi$, where $\psi$ is an $\fo$-formula.

### 2016-08-03 QSO/preliminaries.tex
Nueva versión de QSO.

To define the semantics of $\so$, we need to introduce some terminology. Given a relational signature $\R$ and a finite $\R$-structure $\A$ with domain $A$, a first-order variable-assignment $v$ for $\A$ is a total function from $\fv$ to $A$, while a second-order variable-assignment $V$ for $\A$ is a total function with domain $\sv$ that maps each $X \in \sv$ to a subset of $A^{\arity(X)}$. Moreover, given a first-order variable-assignment $v$ for $\A$, $x \in \fv$ and $a \in A$, we denote by $v[a/x]$ a first-order variable-assignment such that $v[a/x](x) = a$ and $v[a/x](y) = v(y)$ for every $y \in \fv$ distinct from $x$. Similarly, given a second-order variable-assignment $V$ for $\A$, $X \in \sv$ and $B  \subseteq A^{\arity(X)}$, we denote by $V[B/X]$ a second-order variable-assignment such that $V[B/X](X) = B$ and $V[B/X](Y) = V(Y)$ for every $Y \in \sv$ distinct from $X$.

Assume that $\varphi$ is an $\so$-formula over a signature $\R$. Then given a finite $\R$-structure $\A$ with domain $A$, a first-order variable-assignment $v$ for $\A$ and a second-order variable-assignment $V$ for $\A$, we say that $(\A, v, V)$ satisfies $\varphi$, denoted by $(\A, v, V) \models \varphi$, if: (1) $\varphi$ is the formula $R(x_1, \ldots, x_\ell)$ and $(v(x_1), \ldots, v(x_\ell)) \in R^\A$; (2) $\varphi$ is the formula $X(x_1, \ldots, x_m)$ and $(v(x_1), \ldots, v(x_m)) \in V(X)$; (3) $\varphi$ is the formula $\neg \psi$ and it not the case that $(\A, v, V) \models \psi$; (4) $\varphi$ is the formula $(\varphi_1 \vee \varphi_2)$, and $(\A, v, V) \models \varphi_1$ or $(\A, v, V) \models \varphi_2$; (5) $\varphi$ is the formula $\exists x \, \psi$ and there exists $a \in A$ such that $(\A, v[a/x], V) \models \psi$; or (6) $\varphi$ is the formula $\exists X \, \psi$ and there exists $B \subseteq A^{\arity(X)}$ such that $(\A, v, V[B/X]) \models \psi$.  As usual, we consider the propositional operators $\wedge$, $\rightarrow$, and $\leftrightarrow$ that can be obtained from $\vee$ and $\neg$. 
%Moreover, we use the abbreviations $x \not \leq y$ and $x \notin X$ for the negation of the atoms $\leq$ and $\in$. 
%Finally, we consider standard abbreviations of formulas that can be defined in $\so$-logic (actually, $\fo$-logic) like $\first(x) := \fa{x} y \leq x$ and $\last(x) := \fa{x} x \leq y$ to denote the first and last element of the linear order $\leq$, respectively, and $\succesor(x,y) := x \leq y \wedge y \not\leq x \wedge \fa{z} ( z \leq x \vee y \leq z)$ to denote the successor relation.

\subsection{Semirings}

A semiring is a structure $\SR = (S, \adds, \mults, \zero, \one)$ where $(S, \adds, \zero)$ is a commutative monoid, $(S, \mults, \one)$ is a monoid, multiplication ($\mults$) distributes over addition ($\adds$), and $\zero \mults s = s \mults \zero = \zero$ for every $s \in S$. If the multiplication is commutative, we say that $\SR$ is commutative. In this paper, we always assume that $\SR$ is  commutative.
For the sake of simplification, we usually denote the set of elements $S$ with the name of the semiring $\SR$.
% and we say sentences like ``for all $s \in \SR$''.

In general, our definitions are given with respect to a generic semiring.
However, in some cases we consider some particular semiring like, for example, the semiring of \emph{natural numbers} $\nat = (\mathbb{N}, +, \cdot, 0, 1)$, the semiring of \emph{integers} $\integ = (\mathbb{Z}, +, \cdot, 0, 1)$, the \emph{boolean} semiring $\bln = (\{\false, \true\}, \vee, \wedge, \false, \true)$, or the \emph{tropical} semiring $\trop = (\nat \cup \{\infty\}, \min, +, \infty, 0)$.
For the last semiring, the $\min$-operator is defined by the linear order $0 < 1 < 2 < \cdots < \infty$ and the $+$-operator is just the addition over natural number with the exception that $v + \infty = \infty + v = \infty$ for all $v \in \nat \cup \{\infty\}$.

\subsection{Function complexity classes}
Assume that $\R = \{R_1, \ldots, R_k\}$ is a relational signature and $\A$ is a finite $\R$-structure with a domain $A$ containing $n$ elements, and assume that  $<$ is a linear order on $A$, say $a_1 < a_2 < \ldots < a_n$. For every $i \in \{1, \ldots, k\}$, define the encoding of $R_i^\A$, denoted by $\enc(R_i^\A)$, as the following binary string. Assume that $\ell = \arity(R_i)$ and consider an enumeration of the $\ell$-tuples over $A$ in the lexicographic order induced by $<$ (that is, $(a_1, \ldots, a_1, a_1)$, $(a_1, \ldots, a_1, a_2)$, $\ldots$, $(a_n, \ldots, a_n, a_{n-1})$, $(a_n, \ldots, a_n, a_n)$). Then let $\enc(R_i^\A)$ be a binary string of length $n^\ell$ such that the $i$-th bit of $\enc(R_i^\A)$ is 1 if the $i$-th tuple in the previous enumeration belongs to $R_i^\A$, and 0 otherwise. Moreover, define the encoding of $\A$, denoted by $\enc(\A)$, as the following binary string~\cite{L04}:
\begin{eqnarray*}
\enc(\A) & = & 0^n \, 1 \, \enc(R_1^\A) \, \cdots \, \enc(R_k^\A).
\end{eqnarray*}
Given a semiring $\SR = (S, \adds, \mults, \zero, \one)$ and a relational signature $\R$, a function $f$ is said to be from $\R$ to $\SR$ if $f$ is a total function from $\{\enc(\A) \mid \A \in \str[\R]\}$ to $S$. Moreover, a set $\CC$ of functions is said to be a {\em function complexity class over $\SR$} if every $f \in \CC$ is a function from some relational signature $\R$ to $\SR$.

>>>>

To define the semantics of $\so$, we need to introduce some terminology. Given a relational signature $\R$ and a finite $\R$-structure $\A$ with domain $A$, a first-order assignment $v$ for $\A$ is a total function from $\fv$ to $A$, while a second-order assignment $V$ for $\A$ is a total function with domain $\sv$ that maps each $X \in \sv$ to a subset of $A^{\arity(X)}$. Moreover, given a first-order assignment $v$ for $\A$, $x \in \fv$ and $a \in A$, we denote by $v[a/x]$ a first-order assignment such that $v[a/x](x) = a$ and $v[a/x](y) = v(y)$ for every $y \in \fv$ distinct from $x$. Similarly, given a second-order assignment $V$ for $\A$, $X \in \sv$ and $B  \subseteq A^{\arity(X)}$, we denote by $V[B/X]$ a second-order assignment such that $V[B/X](X) = B$ and $V[B/X](Y) = V(Y)$ for every $Y \in \sv$ distinct from $X$.

Assume that $\varphi$ is an $\so$-formula over a signature $\R$. Then given a finite $\R$-structure $\A$ with domain $A$, a first-order assignment $v$ for $\A$ and a second-order assignment $V$ for $\A$, we say that $(\A, v, V)$ satisfies $\varphi$, denoted by $(\A, v, V) \models \varphi$, if: (1) $\varphi$ is the formula $R(x_1, \ldots, x_\ell)$ and $(v(x_1), \ldots, v(x_\ell)) \in R^\A$; (2) $\varphi$ is the formula $X(x_1, \ldots, x_m)$ and $(v(x_1), \ldots, v(x_m)) \in V(X)$; (3) $\varphi$ is the formula $\neg \psi$ and it not the case that $(\A, v, V) \models \psi$; (4) $\varphi$ is the formula $(\varphi_1 \vee \varphi_2)$, and $(\A, v, V) \models \varphi_1$ or $(\A, v, V) \models \varphi_2$; (5) $\varphi$ is the formula $\exists x \, \psi$ and there exists $a \in A$ such that $(\A, v[a/x], V) \models \psi$; or (6) $\varphi$ is the formula $\exists X \, \psi$ and there exists $B \subseteq A^{\arity(X)}$ such that $(\A, v, V[B/X]) \models \psi$.  As usual, we consider the propositional operators $\wedge$, $\rightarrow$, and $\leftrightarrow$ that can be obtained from $\vee$ and $\neg$. 
%Moreover, we use the abbreviations $x \not \leq y$ and $x \notin X$ for the negation of the atoms $\leq$ and $\in$. 
%Finally, we consider standard abbreviations of formulas that can be defined in $\so$-logic (actually, $\fo$-logic) like $\first(x) := \fa{x} y \leq x$ and $\last(x) := \fa{x} x \leq y$ to denote the first and last element of the linear order $\leq$, respectively, and $\succesor(x,y) := x \leq y \wedge y \not\leq x \wedge \fa{z} ( z \leq x \vee y \leq z)$ to denote the successor relation.

\subsection{Function complexity classes}
Assume that $\R = \{R_1, \ldots, R_k\}$ is a relational signature and $\A$ is a finite $\R$-structure with a domain $A$ containing $n$ elements, and assume that  $<$ is a linear order on $A$, say $a_1 < a_2 < \ldots < a_n$. For every $i \in \{1, \ldots, k\}$, define the encoding of $R_i^\A$, denoted by $\enc(R_i^\A)$, as the following binary string. Assume that $\ell = \arity(R_i)$ and consider an enumeration of the $\ell$-tuples over $A$ in the lexicographic order induced by $<$ (that is, $(a_1, \ldots, a_1, a_1)$, $(a_1, \ldots, a_1, a_2)$, $\ldots$, $(a_n, \ldots, a_n, a_{n-1})$, $(a_n, \ldots, a_n, a_n)$). Then let $\enc(R_i^\A)$ be a binary string of length $n^\ell$ such that the $i$-th bit of $\enc(R_i^\A)$ is 1 if the $i$-th tuple in the previous enumeration belongs to $R_i^\A$, and 0 otherwise. Moreover, define the encoding of $\A$, denoted by $\enc(\A)$, as the following binary string~\cite{L04}:
\begin{eqnarray*}
\enc(\A) & = & 0^n \, 1 \, \enc(R_1^\A) \, \cdots \, \enc(R_k^\A).
\end{eqnarray*}
Given a set $\bbD$ and a relational signature $\R$, a function $f$ is said to be from $\R$ to $\bbD$ if $f$ is a total function from $\{\enc(\A) \mid \A \in \str[\R]\}$ to $\bbD$. Moreover, a set $\CC$ of functions over $\bbD$ is said to be a {\em function complexity class over $\bbD$} if every $f \in \CC$ is a function from some relational signature $\R$ to $\bbD$.
